---
title: "data_exploration"
author: "Lauren"
date: "2023-11-20"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## BMEG 423/523 CDSS Assignment
### Initial Data Exploration
Team 2  
2023-11-20  
 
The training and test data used in this project are from the 2019 PhysioNet Computing in Cardiology Challenge. More information about the challenge can be found [here](https://physionet.org/content/challenge-2019/1.0.0/).
 
This document includes the following:  
1. R code to merge all training cases into a single dataframe (which can be easily exported to a csv)  
2. Plots illustrating the characteristics of the dataset  
   b. Comparing patients who never had sepsis to those who didn't have sepsis initially, but developed it, and those who had sepsis from the beginning of the dataset  
   c. Characteristics of patient data 4-6 hours prior to sepsis diagnosis  
  
```{r}
#LOAD REQUIRED PACKAGES
library(tidyverse)
```
  
#### Merge the raw training and test data and save as an RData file
```{r}
#FUNCTION TO MERGE ALL TRAINING CASES INTO ONE DATAFRAME AND SAVE AS RData FILE AND MERGE ALL TEST CASES INTO ANOTHER DATAFRAME # This code is based on code provided in the assignment package
load_cinc_data <- function(fromfile = T){
  #IF THE MERGED DATA FILE ALREADY EXISTS, JUST LOAD IT IN
  if (fromfile == T){
    load("CinC.RData")
  } 
  
  else {
    #CREATE A FILE CALLED CINC.Rdata TO HOLD THE MERGED TRAINING DATA FROM ALL PATIENTS WITH A COLUMN ADDED FOR PATIENT ID
    files <- list.files(here::here("training_2023-11-05"), full.names = TRUE) #creates character vector of the names of files in the training data folder
    cinc_dat <- NULL #initialize data frame to hold all patient data
    for (f in files){
      fname <- substr(basename(f), 1, nchar(basename(f))-4) #basically just removes ".csv" from the file names
      #print(fname)
      pdat <- read.delim(f, sep = ",", na = "NA") #create temp variable called pdat that contains a data frame read in from the patient data csv file
      pdat <- cbind(patient = fname, pdat) #adds a column to the data frame with the patient ID
      cinc_dat <- rbind(cinc_dat, pdat) #add data for current patient to data frame with all patient data (bind by row)
    }
    # Save the data
    save(cinc_dat, file = "CinC.RData")
    
    #CREATE A FILE CALLED CINC_Test.Rdata TO HOLD THE MERGED TEST DATA FROM ALL PATIENTS WITH A COLUMN ADDED FOR PATIENT ID
    files <- list.files(here::here("testing_2023-11-05"), full.names = TRUE)
    cinc_dat_test <- NULL
    for (f in files){
      fname <- substr(basename(f), 1, nchar(basename(f))-4)
      #print(fname)
      pdat <- read.delim(f, sep = ",", na = "NA")
      pdat <- cbind(patient = fname, pdat)
      cinc_dat_test <- rbind(cinc_dat_test, pdat)
    }
    # Save the data
    save(cinc_dat_test, file = "CinC_Test.RData")
  }
  return(cinc_dat)
}
```

#### Load the merged data to a data frame in the working environment
```{r}
#cinc_data <- load_cinc_data(fromfile = F) #run this to merge data and save RData files if they don't exist yet
cinc_data <- load_cinc_data() #run this to load merged training data into a data frame called cinc_data if the RData files already exist
```
  
#### Organize the data
```{r}
#Get vector with all patient IDs
patientIDs <- cinc_data %>%
  select(patient) %>%
  distinct(patient)

#Get patient IDs for all patients with a positive sepsis label at some point
sepsis_patientIDs <- cinc_data %>%
  filter(SepsisLabel == 1) %>%
  distinct(patient)

#Get patient IDs for all patients who never developed sepsis
no_sepsis_patientIDs <- anti_join(patientIDs, sepsis_patientIDs, by = "patient")

#Data frame with all data for patients with a positive sepsis label at some point
sepsis <- cinc_data %>%
  filter(patient %in% sepsis_patientIDs[[1]])

#Data frame with all data for patients who never developed sepsis
no_sepsis <- cinc_data %>%
  filter(patient %in% no_sepsis_patientIDs[[1]])
```

#### Start analyzing data
[note to self - add in means for 6 and 4 hours before positive sepsis diagnosis]
```{r}
# Means computed for all sepsis-positive patient data (including data before and after positive sepsis diagnosis)
# NA values have been removed
sepsis_means <- sepsis %>%
  summarise(mean_HR = mean(HR, na.rm = TRUE),
            mean_O2sat = mean(O2Sat, na.rm = TRUE),
            mean_Temp = mean(Temp, na.rm = TRUE),
            mean_SBP = mean(SBP, na.rm = TRUE),
            mean_MAP = mean(MAP, na.rm = TRUE),
            mean_DBP = mean(DBP, na.rm = TRUE),
            mean_Resp = mean(Resp, na.rm = TRUE),
            mean_FiO2 = mean(FiO2, na.rm = TRUE),
            mean_pH = mean(pH, na.rm = TRUE),
            mean_BUN = mean(BUN, na.rm = TRUE),
            mean_Creatinine = mean(Creatinine, na.rm = TRUE),
            mean_Glucose = mean(Glucose, na.rm = TRUE),
            mean_Magnesium = mean(Magnesium, na.rm = TRUE),
            mean_Potassium = mean(Potassium, na.rm = TRUE),
            mean_Hct = mean(Hct, na.rm = TRUE),
            mean_Hgb = mean(Hgb, na.rm = TRUE),
            mean_WBC = mean(WBC, na.rm = TRUE),
            mean_Age = mean(Age, na.rm = TRUE)) %>%
  mutate(Demographic = "All data for sepsis-positive patients before and after positive SepsisLabel") %>%
  select(Demographic, everything())

# Means computed for sepsis-positive patient data (only including data during which SepsisLabel = 1)
# NA values have been removed
sepsis_means_after <- sepsis %>%
    filter(SepsisLabel == 1) %>%
    summarise(mean_HR = mean(HR, na.rm = TRUE),
            mean_O2sat = mean(O2Sat, na.rm = TRUE),
            mean_Temp = mean(Temp, na.rm = TRUE),
            mean_SBP = mean(SBP, na.rm = TRUE),
            mean_MAP = mean(MAP, na.rm = TRUE),
            mean_DBP = mean(DBP, na.rm = TRUE),
            mean_Resp = mean(Resp, na.rm = TRUE),
            mean_FiO2 = mean(FiO2, na.rm = TRUE),
            mean_pH = mean(pH, na.rm = TRUE),
            mean_BUN = mean(BUN, na.rm = TRUE),
            mean_Creatinine = mean(Creatinine, na.rm = TRUE),
            mean_Glucose = mean(Glucose, na.rm = TRUE),
            mean_Magnesium = mean(Magnesium, na.rm = TRUE),
            mean_Potassium = mean(Potassium, na.rm = TRUE),
            mean_Hct = mean(Hct, na.rm = TRUE),
            mean_Hgb = mean(Hgb, na.rm = TRUE),
            mean_WBC = mean(WBC, na.rm = TRUE),
            mean_Age = mean(Age, na.rm = TRUE)) %>%
  mutate(Demographic = "Data for sepsis-positive patients after positive SepsisLabel") %>%
  select(Demographic, everything())

# Means computed for sepsis-positive patient data (including only data before positive sepsis diagnosis)
# NA values have been removed
sepsis_means_before <- sepsis %>%
  filter(SepsisLabel == 0) %>%
  summarise(mean_HR = mean(HR, na.rm = TRUE),
            mean_O2sat = mean(O2Sat, na.rm = TRUE),
            mean_Temp = mean(Temp, na.rm = TRUE),
            mean_SBP = mean(SBP, na.rm = TRUE),
            mean_MAP = mean(MAP, na.rm = TRUE),
            mean_DBP = mean(DBP, na.rm = TRUE),
            mean_Resp = mean(Resp, na.rm = TRUE),
            mean_FiO2 = mean(FiO2, na.rm = TRUE),
            mean_pH = mean(pH, na.rm = TRUE),
            mean_BUN = mean(BUN, na.rm = TRUE),
            mean_Creatinine = mean(Creatinine, na.rm = TRUE),
            mean_Glucose = mean(Glucose, na.rm = TRUE),
            mean_Magnesium = mean(Magnesium, na.rm = TRUE),
            mean_Potassium = mean(Potassium, na.rm = TRUE),
            mean_Hct = mean(Hct, na.rm = TRUE),
            mean_Hgb = mean(Hgb, na.rm = TRUE),
            mean_WBC = mean(WBC, na.rm = TRUE),
            mean_Age = mean(Age, na.rm = TRUE)) %>%
  mutate(Demographic = "Data for sepsis-positive patients before positive SepsisLabel") %>%
  select(Demographic, everything())

# Means computed for all sepsis-negative patient data (only including patients who never developed sepsis)
# NA values have been removed
no_sepsis_means <- no_sepsis %>%
  summarise(mean_HR = mean(HR, na.rm = TRUE),
            mean_O2sat = mean(O2Sat, na.rm = TRUE),
            mean_Temp = mean(Temp, na.rm = TRUE),
            mean_SBP = mean(SBP, na.rm = TRUE),
            mean_MAP = mean(MAP, na.rm = TRUE),
            mean_DBP = mean(DBP, na.rm = TRUE),
            mean_Resp = mean(Resp, na.rm = TRUE),
            mean_FiO2 = mean(FiO2, na.rm = TRUE),
            mean_pH = mean(pH, na.rm = TRUE),
            mean_BUN = mean(BUN, na.rm = TRUE),
            mean_Creatinine = mean(Creatinine, na.rm = TRUE),
            mean_Glucose = mean(Glucose, na.rm = TRUE),
            mean_Magnesium = mean(Magnesium, na.rm = TRUE),
            mean_Potassium = mean(Potassium, na.rm = TRUE),
            mean_Hct = mean(Hct, na.rm = TRUE),
            mean_Hgb = mean(Hgb, na.rm = TRUE),
            mean_WBC = mean(WBC, na.rm = TRUE),
            mean_Age = mean(Age, na.rm = TRUE)) %>%
  mutate(Demographic = "All data for sepsis-negative patients") %>%
  select(Demographic, everything())

means <- rbind(sepsis_means, sepsis_means_after, sepsis_means_before, no_sepsis_means)
print(means)
```


